\erratumItem{4.{}2}{wrong reference}
\erratumItem{4.{}2}{wrong cross-reference for ``line 16''}
\erratumItem{4.{}3}{{\texttt  {for}} attribute on {\texttt  {definition}} should be of type {\texttt  {NCNames}}}
\erratumItem{4.{}3}{should be "definiendum" not "definiens"}
\erratumItem{4.{}3}{should be definiendum-applied not definiens-applied}
\erratumItem{4.{}4}{{\texttt  {for}} attribute on {\texttt  {definition}} should be of type {\texttt  {NCNames}}}
\erratumItem{4.{}4}{{\texttt  {for}} attribute on {\texttt  {definition}} should be of type {\texttt  {NCNames}}}
\erratumItem{4.{}4}{should be "definiendum" not "definiens"}
\erratumItem{5.}{{\texttt  {for}} attribute on {\texttt  {definition}} should be of type {\texttt  {NCNames}}}
\erratumItem{6.}{{\texttt  {for}} attribute on {\texttt  {definition}} should be of type {\texttt  {NCNames}}}
\erratumItem{6.}{{\texttt  {for}} attribute on {\texttt  {definition}} should be of type {\texttt  {NCNames}}}
\erratumItem{7.}{{\texttt  {for}} attribute on {\texttt  {definition}} should be of type {\texttt  {NCNames}}}
\erratumItem{7.}{{\texttt  {for}} attribute on {\texttt  {definition}} should be of type {\texttt  {NCNames}}}
\erratumItem{7.}{{\texttt  {for}} attribute on {\texttt  {definition}} should be of type {\texttt  {NCNames}}, totally reworked example}
\erratumItem{8.{}1}{{\texttt  {for}} attribute on {\texttt  {axiom}} should be of type {\texttt  {NCNames}}}
\erratumItem{8.{}1}{{\texttt  {for}} attribute on {\texttt  {definition}} should be of type {\texttt  {NCNames}}}
\erratumItem{8.{}1}{forgot to thread through attribute renaming}
\erratumItem{8.{}1}{The attribute on the {\tt  {assertion}} element should be {\tt  {just-by}}, not {\tt  {proofs}}. We were also missing some fragment identifiers.}
\erratumItem{11.{}1}{Typo: ``Backus Naur form'' instead of ``Bachus Naur Form''}
\erratumItem{11.{}1}{ref does permit an xml:id attribute (and this should remain, as that is important for talking about refs from an RDF point of view)}
\erratumItem{11.{}1}{omdoc and omgroup can have an optional theory attribute as well}
\erratumItem{11.{}2}{RDF as a general data model is independent from XML; RDF/XML is just one of its possible serializations.}
\erratumItem{11.{}2}{correct name}
\erratumItem{12.}{The content Model for {\tt  {dc:creator}} and {\tt  {cd:contributor}} is simple text}
\erratumItem{12.{}1}{wrong attribute name}
\erratumItem{12.{}4}{{\texttt  {for}} attribute on {\texttt  {definition}} should be of type {\texttt  {NCNames}}}
\erratumItem{13.{}1.1}{It should be made clear that this inheritance mechanism is extended by the OMDoc format. See section {\textbf  {3.1}} of the errata document for details}
\erratumItem{14.}{added the attribute xml:id to the {\tt  CMP} element; added the attribute from to the {\tt  omtext} element}
\erratumItem{14.}{added the attribute cdbase to the {\tt  term} element}
\erratumItem{14.{}1}{should be "definiendum" not "definiens"}
\erratumItem{14.{}3}{\attval {note}{type}{omtext}}
\erratumItem{14.{}3}{omtext can also be an assumption, obligation or rule as all of these can be expressed in informal as well as formal way}
\erratumItem{14.{}3}{And there should also be <omtext type=``assertion''> for generic assertions, corresponding to the <assertion> element without a type.}
\erratumItem{14.{}5}{also need {\tt  cdbase} for identifying}
\erratumItem{14.{}5}{Should be Deffiniendum instead of deffiniens}
\erratumItem{14.{}5}{should be "definiendum" not "definiens"}
\erratumItem{14.{}6}{the index attribute should be optional}
\erratumItem{15.{}1}{``Definiendum'' and ``Definiens'' should switched}
\erratumItem{15.{}2.1}{scope is deprecated}
\erratumItem{15.{}2.2}{the {\texttt  {for}} attribute in the {\texttt  {axiom}} element must reference symbol names}
\erratumItem{15.{}2.2}{{\texttt  {for}} attribute on {\texttt  {axiom}} should be of type {\texttt  {NCNames}}}
\erratumItem{15.{}2.3}{examples reference wrong listings}
\erratumItem{15.{}2.4}{Note that this use of the {\texttt  {for}} attribute is different from the other usages, which are URI references.}
\erratumItem{15.{}2.4}{{\texttt  {for}} attribute on {\texttt  {definition}} should be of type {\texttt  {NCNames}}, also corrected cd attribute.}
\erratumItem{15.{}2.4}{{\texttt  {for}} attribute on {\texttt  {definition}} should be of type {\texttt  {NCNames}}}
\erratumItem{15.{}2.4}{{\texttt  {for}} attribute on {\texttt  {definition}} should be of type {\texttt  {NCNames}}}
\erratumItem{15.{}3}{deleted spurious {\tt  {for}} attribute on the {\tt  {assertion}} element, {\texttt  {alternative}} should have the same content as definition}
\erratumItem{15.{}3.2}{{\texttt  {for}} attribute on {\texttt  {definition}} should be of type {\texttt  {NCNames}}}
\erratumItem{15.{}3.3}{fixed the target of the for attribute}
\erratumItem{15.{}4}{added the {\tt  axiom} element to the list; cf.\ discussion on {\tt  omdoc-dev} on May 16, 2008}
\erratumItem{15.{}4}{added the {\tt  alternative} element to the list}
\erratumItem{15.{}4}{{\texttt  {for}} attribute on {\texttt  {definition}} should be of type {\texttt  {NCNames}}}
\erratumItem{15.{}5}{{\texttt  {for}} attribute on {\texttt  {definition}} should be of type {\texttt  {NCNames}}}
\erratumItem{15.{}5}{should be "definendum" not "definiens"}
\erratumItem{15.{}5}{{\texttt  {for}} attribute on {\texttt  {definition}} should be of type {\texttt  {NCNames}}}
\erratumItem{15.{}5}{should be "definiendum" not "definiens"}
\erratumItem{15.{}6}{the {\tt  {xml:id}} attribute on the {\tt  {theory}} element should be optional}
\erratumItem{15.{}6.1}{The symbol name {\texttt  {af}} should be {\texttt  {aa}}}
\erratumItem{15.{}6.1}{{\texttt  {for}} attribute on {\texttt  {definition}} should be of type {\texttt  {NCNames}}}
\erratumItem{15.{}6.1}{{\texttt  {for}} attribute on {\texttt  {definition}} should be of type {\texttt  {NCNames}}}
\erratumItem{15.{}6.2}{This specification of the inheritance mechanism is to wishy washy. See section {\textbf  {3.1}} of the errata document for a clarification.}
\erratumItem{16.}{The {\tt  {for}} attribute contains a URI reference according to the RelaxNG schema; the locality restriction here contradicts that and needs to be removed.}
\erratumItem{17.{}1}{{\texttt  {for}} attribute on {\texttt  {definition}} should be of type {\texttt  {NCNames}}}
\erratumItem{17.{}1}{made the for attribute in the {\tt  proofobject} element required; added the rank attribute to the {\tt  premise} element}
\erratumItem{17.{}2}{{\texttt  {for}} attribute on {\texttt  {definition}} should be of type {\texttt  {NCNames}}}
\erratumItem{17.{}2}{{\texttt  {for}} attribute on {\texttt  {definition}} should be of type {\texttt  {NCNames}}}
\erratumItem{17.{}3}{{\texttt  {for}} attribute on {\texttt  {definition}} should be of type {\texttt  {NCNames}}}
\erratumItem{17.{}4}{{\texttt  {for}} attribute on {\texttt  {definition}} should be of type {\texttt  {NCNames}}}
\erratumItem{18.}{changed the order of type and hiding attributes in the {\tt  morphism} element; removed the consistency and consistency-just attributes from the {\tt  morphism}, {\tt  inclusion}, {\tt  theory-inclusion}, and {\tt  axiom-inclusion} elements; changed the contents of the {\tt  theory-inclusion} element to (morphism?, obligation*); changed the contents of the {\tt  morphism} element to (requation+, measure?, ordering?); added the element {\tt  obligation}}
\erratumItem{18.{}1}{noted special case}
\erratumItem{18.{}1}{Clarified wording}
\erratumItem{18.{}2}{added missing word}
\erratumItem{18.{}2}{Fixed value of the {\tt  conservativity} attribute}
\erratumItem{18.{}2}{Fixed value of the {\tt  conservativity} attribute}
\erratumItem{18.{}2}{Fixed value of the {\tt  conservativity} attribute}
\erratumItem{18.{}5.2}{added the optional for attribute for the {\\decomposition} element; removed the by attribute from the {\tt  theory-inclusion} element; changed the contents of the {\tt  theory-inclusion} element to (morphism?, (decomposition* | obligation*))}
\erratumItem{19.}{added CMP* to content of {\tt  presentation} element}
\erratumItem{19.{}4}{The {\texttt  {for}} attribute should be {\texttt  {\#X4}} instead of {\texttt  {\#X}} in listings 19.5 and 19.6}
\erratumItem{20.{}1}{The reference {\tt  {reformulates="ALGX0"}} should be a URI reference, i.e. {\tt  {\#ALGX0}}}
\erratumItem{20.{}2}{Wrong Content Model for {\tt  {omlet}}}
\erratumItem{22.{}1}{{\texttt  {for}} attribute on {\texttt  {definition}} should be of type {\texttt  {NCNames}}}
\erratumItem{22.{}1}{{\texttt  {for}} attribute on {\texttt  {definition}} should be of type {\texttt  {NCNames}}}
\erratumItem{22.{}2}{{\texttt  {for}} attribute on {\texttt  {definition}} should be of type {\texttt  {NCNames}}}
\erratumItem{26.{}4}{reference to QED}
\erratumItem{26.{}15}{The domain is \texttt  {kwarc.eecs.iu-bremen.de}}
\erratumItem{26.{}15.4}{correct example given}
\erratumItem{1.{}1}{the old \texttt  {extradata} content has nothing to do with \texttt  {dc:subject}}
\erratumItem{4.{}4}{The {\texttt  {type}} attributes on {\texttt  {phrase}} and {\texttt  {omtext}} were not conforming to the spec}
\erratumItem{4.{}4}{attribute value {\texttt  {trasition}} forgotten from rnc}
\erratumItem{4.{}4}{the {\texttt  {verbalizes}} attribute had been forgotten for the {\texttt  {phrase}} element}
\erratumItem{4.{}7}{simple definitions should not have an {\tt  {existence}} attribute, furthermore pattern definitions should not have {\texttt  {measure}} and {\texttt  {ordering}} children}
\erratumItem{4.{}7}{the {\tt  {type}} element needs to allow a {\tt  {for}} attribute}
\erratumItem{4.{}7}{the {\texttt  {tgroup}} element should not contain {\texttt  {omgroup}} children}
\erratumItem{4.{}10}{The DG module RelaxNG schema had been forgotten}
\erratumItem{4.{}13}{we have to allow the {\tt  {metadata}} element in {\tt  {omlet}}}
